---
title: "Functions - Pre-Class Work"
author: "Ozan Adiguzel (Override)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Standardizing a variable means subtracting the mean, and then dividing by the standard deviation. Let’s use a loop to standardize the numeric columns in the [Western Collaborative Group Study](https://clinicaltrials.gov/ct2/show/NCT00005174). This study began in 1960 with 3154 men ages 39-59, who were employed in one of 11 California based companies. They were followed until 1969 during this time, 257 of these men developed coronary heart disease (CHD). You can read this data in with the code below. You can access this dataset with the following code:

```{R}

suppressMessages(library(foreign))
# downloaded the data since the link didn't work
# although total number of observations is same, some variable names differ from the table below so the downloaded dataset (wcgs2) may be slightly  different than the originally implemented one
wcgs <- read.dta("../wcgs2.dta")
```

The data has the following variables:

WCGS has the following variables:

-----------------------------------------------------------
Name    Description
------- -------------------------------------------
id      Subject identification number

age     Age in years

height  Height in inches

weight  Weight in lbs.

sbp     Systolic blood pressure in mm 

dbp     Diastolic blood pressure in mm Hg

chol    Fasting serum cholesterol in mm 

behpat  Behavior

  1       A1

  2       A2

  3       B3

  4       B4

ncigs   Cigarettes per day

dibpat  Behavior

1       type A

2       type B

chd69   Coronary heart disease

1       Yes

0       no

typechd Type of CHD

1       myocardial infarction or death

2       silent myocardial infarction

3       angina perctoris

time169 Time of CHD event or end of follow-up

arcus   Arcus senilis

0       absent

1       present

bmi     Body Mass Index
-----------------------------------------------------------

### Question 1: Standardize Function

A. Create a function called standardize.me() that takes a numeric vector as an argument, and returns the standardized version of the vector. 

```{r}
standardize.me <- function(x) {
  (x - mean(x)) / sd(x)
}
```

B. Assign all the numeric columns of the original WCGS dataset to a new dataset called WCGS.new.

  This operation can be done in different ways as I show below. However, I'd prefer to use map_lgl function rather than sapply function. Because all functions in the purr package are type-consistent while sapply is type-inconsistent.

```{r}
# install.packages("purrr")
library(purrr)
nmrc <- map_lgl(wcgs, is.numeric)
WCGS.new <- wcgs[ , nmrc]
str(WCGS.new)
```

```{r}
nmrc <- sapply(wcgs, is.numeric)
WCGS.new <- wcgs[ , nmrc]
str(WCGS.new)
```

C. Using a loop and your new function, standardize all the variables WCGS.new dataset.

```{r}
for(p in seq_along(WCGS.new)) {
  WCGS.new[ , p] <- standardize.me(WCGS.new[ , p])
}
head(WCGS.new)
```

D. What should the mean and standard deviation of all your new standardized variables be? Test your prediction by running a loop

  Since we standardized our variables, they should all have average of 0 and standard deviation of 1. I confirmed this prediction by running a loop as well as using map and sapply functions below. (For all three methods below; na.rm=T argument is optiona since only "chol" includes NAs, but it is a good practice for the flexibility of the code)

```{r}
for(p in seq_along(WCGS.new)) {
  print(
    paste(
      "For column ", p, ", average: ", round(mean(WCGS.new[,p], na.rm = T), 3), 
      " - standard deviation: ",round(sd(WCGS.new[,p], na.rm = T), 3)
    )
  )
}
```

```{r}
map(WCGS.new, function(col) list(mean = round(mean(col, na.rm = T), 3), sd = round(sd(col, na.rm = T), 3)))
```

```{r}
sapply(WCGS.new, function(col) list(mean = round(mean(col, na.rm = T), 3), sd = round(sd(col, na.rm = T), 3)))
```


### Question 2: Looping to Calculate

A. Using a loop, calculate the mean weight of the subjects separated by the type of CHD they have.

```{r}
# create a dataframe to fill in
chdAvgWeights <- data.frame("CHDtypes" = unique(wcgs$typchd69),
                        "AvgWeights" = NA
                        )
# fill in the data frame by subsetting the unique CHD types and calculating the corresponding mean values
for(n in 1:nrow(chdAvgWeights)) {
 CHD <- chdAvgWeights$CHDtypes[n]
 Avg <- mean(subset(wcgs, typchd69 == CHD)$weight, na.rm = T)
 chdAvgWeights$AvgWeights[n] <- Avg
}
# print the modified dataframe
chdAvgWeights
```

B. Now do the same thing, but now don’t use a loop

The same operation could be done with aggregate function or with a combination of dplyr functions as shown below. Actually, looping is an inefficient option compared to these other methods.

```{r}
# install.packages("dplyr")
library("dplyr")
wcgs %>% group_by(typchd69) %>%
  summarise(avgWeights = mean(weight, na.rm = T))
```

```{r}
aggregate(weight ~ typchd69, 
          data = wcgs, 
          FUN = mean)
```



